Raw Loan Analysis
========================

### Load libraries

```{r}
#library(RColorBrewer)
```

  
## Processing

Download the data, read the data in and save the raw data as an RDA file

```{r loadData,cache=TRUE}
getwd()
download.file("https://spark-public.s3.amazonaws.com/dataanalysis/loansData.rda", destfile="../../data/loans.rda",method="curl")
date()
load("../../data/loans.rda")
loansRaw <- loansData
```


### Look at the data set

```{r, dependson="loadData"}
head(loansRaw)
summary(loansRaw)
sapply(loansRaw[1,],class)
```

### Find out about missing values
```{r, dependson="loadData"}
sum(is.na(loansRaw)) 
```

### Find minimum and maximum interest rates
```{r, dependson="loadData"}
rates <- as.numeric(sub("%", "", loansRaw$Interest.Rate))
min(rates)
max(rates)
```

--------------------------

## Exploratory analysis

### Make some univariate plots/summaries


```{r, dependson="loadData"}
hist(rates)
hist(log(loansRaw$Monthly.Income), breaks=50) 
plot(loansRaw$FICO.Range, las=2) 
hist(loansRaw$Amount.Funded.By.Investors)
hist(loansRaw$Open.CREDIT.Lines)
hist(log(loansRaw$Revolving.CREDIT.Balance), breaks=50)
plot(loansRaw$Employment.Length, las=2)
```

### Make some univariate tables
```{r, dependson="loadData"}
table(loansRaw$FICO.Range)
table(loansRaw$Interest.Rate)
table(loansRaw$Loan.Length)
table(loansRaw$State)
table(loansRaw$Loan.Purpose)
table(loansRaw$Home.Ownership)
table(loansRaw$Inquiries.in.the.Last.6.Months)
table(loansRaw$Employment.Length)
```

### Plot rate by FICO
```{r, dependson="loadData"}
plot(loansRaw$FICO.Range, rates,pch=19, las=2)
```

Have correlation between FICO and Rate.
Than bigger FICO than less Rate. 
FICO is itslef aggregation of other factors.

### Plot Rate by FICO with color groups for loan lengths 
```{r, dependson="loadData"}
plot(as.numeric(loansRaw$FICO.Range), rates, pch=19, col=loansRaw$Loan.Length)
```
Rates smaller than loan length smaller.

### Plot rate by purpose
```{r, dependson="loadData"}
plot(loansRaw$Loan.Purpose, rates, las=2)
```
In general car, educational and energy has lower rate, maby because they are short term and then loan length is cofounder here


### Plot rate by Inquiries.in.the.Last.6.Months 
```{r, dependson="loadData"}
plot(loansRaw$Inquiries.in.the.Last.6.Months, rates, col=loansRaw$Loan.Length)
```
Than more quantity of inquires than more rate (at least this works after 3d inquiry)


------

## Modeling

### Fit a rates by FICO.Range
```{r, dependson="loadData"}
lm1 <- lm(rates ~loansRaw$FICO.Range)
plot(loansRaw$FICO.Range, rates, las=2)
points(loansRaw$FICO.Range, lm1$fitted,pch=19,col="red")
```

###Fit a rates by loansRaw$Amount.Funded.By.Investors
```{r, dependson="loadData"}
lm4 <- lm(rates ~ loansRaw$Amount.Funded.By.Investors)
plot(loansRaw$Amount.Funded.By.Investors, rates)
points(loansRaw$Amount.Funded.By.Investors, lm4$fitted,pch=19,col="red")
```

###Fit a rates by Open.CREDIT.Lines
```{r, dependson="loadData"}
rates1 <- rates[!is.na(loansRaw$Open.CREDIT.Lines)]
lines <- loansRaw$Open.CREDIT.Lines[!is.na(loansRaw$Open.CREDIT.Lines)]
lm10 <- lm(rates1 ~ lines)
plot(lines, rates1, las=2)
points(lines, lm10$fitted,pch=19,col="red")
```

###Fit a rates by Inquiries.in.the.Last.6.Months
```{r, dependson="loadData"}
rates1 <- rates[!is.na(loansRaw$Inquiries.in.the.Last.6.Months)]
inquires <- loansRaw$Inquiries.in.the.Last.6.Months[!is.na(loansRaw$Inquiries.in.the.Last.6.Months)]
lm11 <- lm(rates1 ~ inquires)
plot(inquires, rates1)
points(inquires, lm11$fitted,pch=19,col="red")
```

## Let's use this model

```{r, dependson="loadData"}
ratio1 <- cut(ratio, 3)
range <- cut(as.numeric(loansRaw$FICO.Range), 5)
lines <- cut(as.numeric(loansRaw$Open.CREDIT.Lines), 3)
lmFinal <- lm(rates ~ loansRaw$Amount.Funded.By.Investors + range + loansRaw$Loan.Length + loansRaw$Inquiries.in.the.Last.6.Months + loansRaw$Home.Ownership)
```

## Get the estimates and confidence intervals

```{r, dependson="lmFinalChunk"}
## The estimate from summary
summary(lmFinal)
## The confidence interval from confint
confint(lmFinal)
```

-------

## Figure making
```{r, dependson="lmFinalChunk"}
## Set up a function that makes colors prettier

## Save figure to pdf file
pdf(file="../../figures/finalfigure.pdf", height=4, width=3*4)

## Set size of axes
cx = 1.3

par(mfrow=c(1,3))

hist(rates,col="yellow", xlab="Rates",ylab="Frequency",main="",cex.axis=cx,cex.lab=cx)
plot(as.numeric(loansRaw$FICO.Range), rates, pch = 19, col = loansRaw$Loan.Length, cex.axis=cx,cex.lab=cx, xlab="FICO Range", ylab="Rate")
plot(loansRaw$Amount.Funded.By.Investors, rates, col=range, pch=ifelse(as.numeric(loansRaw$Loan.Length) == 2, 19, 21), cex.axis=cx,cex.lab=cx, xlab="Amount funded (USD)", ylab="Rate")

dev.off()
```

